Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-03 conclusions.tex
lloid_paper: reworked astrophysics a bit in intro and conclusion
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@924 ccdd05d5-56c1-461a-bf98-85f6876523b4

\editorial{We should drive home the point that our method is as fast as the fft convolution but without any latency at all.}%
We have demonstrated a computationally feasible procedure for the rapid and even advance
detection of gravitational waves emitted during the coalescence of neutron
stars and stellar-mass black holes.  These sources are expected to produce
prompt electromagnetic signals and may be the progenitors of some short hard
gamma-ray bursts.  Rapid alerts to the broader astronomical community may
improve the chances of detecting an electromagnetic counterpart in bands from
X-ray down to radio.  We anticipate requiring $\sim$\numcpus\ modern
%
\editorial{We need to do this calculation using the \flops\ counts and the
number of templates. CHAD: Agreed, but if you do it by the books it seems
misleading.  I have adjusted it to 600.  This number is taken from assuming
that 1 657 template sub-bank can be filtered on one core and dividing the total
number of templates (10$^5$) by 657 which gives you 150 cores per detector. I
put it in section 2. }%
%
computer cores to analyze a four-detector network of gravitational-wave data
for binary neutron stars and stellar mass black holes.  This is within the
current computing capabilities of the \textsc{lsc} Data Grid~\cite{LDG}.

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence introducing an up-front latency.  Data
aggregation from the observatories will travel over various networks, each
capable of high bandwith but perhaps only modest latency.  This could amount to
a similar latency of $\sim$100 ms.  Lastly, unless a realtime infrastructure is
adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation using \gstlal\ that is capable of $\sim$1 s latency. In our
opinion, significant work would have to be done in order to improve upon this
number. However, it should be considered for third-generation detector design.
For example, a tighter integration of analysis and data acquisition would be
beneficial.

>>>>

\editorial{We should drive home the point that our method is as fast as the fft convolution but without any latency at all.}%
We have demonstrated a computationally feasible procedure for the rapid and even preemptive
detection of gravitational waves emitted during the coalescence of neutron
stars and stellar-mass black holes. Our method is as fast as standard \fft{}
convolutions but without any latency at all. Our search targets are expected to
produce prompt electromagnetic signals and may be the progenitors of some short
hard gamma-ray bursts.  Rapid alerts to the broader astronomical community will
improve the chances of detecting an electromagnetic counterpart in bands from
gamma-rays down to radio.  We anticipate requiring $\sim$\numcpus\ modern
%
\editorial{We need to do this calculation using the \flops\ counts and the
number of templates. CHAD: Agreed, but if you do it by the books it seems
misleading.  I have adjusted it to 600.  This number is taken from assuming
that 1 657 template sub-bank can be filtered on one core and dividing the total
number of templates (10$^5$) by 657 which gives you 150 cores per detector. I
put it in section 2. }%
%
computer cores to analyze a four-detector network of gravitational-wave data
for binary neutron stars and stellar-mass black holes.  This is well within the
current computing capabilities of the \textsc{lsc} Data Grid~\cite{LDG}.

The algorithm we described has no intrinsic latency.  However, there are
fundamental and practical latencies associated with the analysis and detection
procedure. For example, the \LIGO{} detectors, data acquisition is synchronized
to a 1/(16~Hz) cadence introducing an up-front latency.  Data
aggregation from the observatories will travel over various networks, each
capable of high bandwith but perhaps only modest latency.  This could amount to
a similar latency of $\sim$100 ms.  Lastly, unless a real-time infrastructure is
adopted post data acquisition, it is likely that there will be an inherent
latency introduced by such infrastructure.  We have shown a prototype
implementation using \gstlal\ that is capable of $\sim$1 s latency. In our
opinion, significant work would have to be done in order to improve upon this
number. For example, a tighter integration of analysis and data acquisition
would be beneficial. This should be considered for third-generation detector
design.

### 2011-06-03 introduction.tex
lloid_paper: reworked astrophysics a bit in intro and conclusion
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@924 ccdd05d5-56c1-461a-bf98-85f6876523b4

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short GRBs)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
GRBs~\cite{1999Natur.398..400A,2003Natur.422..284F,2006A&A...454L.119J,0004-637X-660-1-489,2008Natur.455..183R,2011A&A...528A..15G}
by telescopes with extremely rapid response or, in the case of GRB 080319B, by
pure serendipity, where several telescopes were observing a previous GRB in the
same field~\cite{2008Natur.455..183R}. Short GRBs, on the other hand, typically
fade too quickly to observe anything before the initial burst of gamma rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short GRBs. An optical counterpart would vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting gravitational-wave candidates not minutes
\emph{after} the merger, but seconds \emph{before}.  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for NS--NS binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.

>>>>

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which are
relatively insensitive to initial conditions. Optical flashes
have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A}
by telescopes with extremely rapid response or, in the case of \textsc{grb 080319b}, by
pure serendipity, where several telescopes were observing a previous \GRB{} in the
same field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to observe anything before the initial burst of gamma-rays and hard
x-rays. Rapid gravitational-wave transient alerts could enable the observation
of optical flashes from short \GRB{}s and perhaps even
precursors~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost
confidence in the gravitational-wave detection and provide the tight sky
localization necessary to allow determination of the source's host galaxy,
which leads to a redshift measurement. With both redshift and a coincident
gravitational-wave observation, we can produce precision measurements of the
Hubble constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By
looking for threshold crossings before the gravitational-wave signal leaves
the detection band, it is possible to trade some signal to noise ratio (\SNR{})
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for NS--NS binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

@@

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
principal component analysis to reduce the number of templates.  We described a
rank-reduction scheme based on singular value decomposition in
\cite{Cannon:2010p10398}.  We will use both aspects to demonstrate that a very
low latency analysis with advance detection of compact binary sources is
possible with current computing resources.  Assuming other technical sources of
latency can be reduced significantly, this should allow the possibility for
prompt alerts to be sent to the astronomical community.
